Supplementary Material

### 2024-03-20 article/paper_drought_Chile2023.tex
update article

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

To analyze land cover change, we use the classification scheme by the
IGBP (International Geosphere-Biosphere Programme) from the product
MCD12Q1 collection 6.1 from MODIS. To derive a proxy for vegetation
productivity, we used the Normalized Difference Vegetation Index (NDVI)
from the product MOD13A3 collection 6.1 from MODIS \citep{Didan2015}.
MOD13A3 provides vegetation indices at 1km of spatial resolution and
monthly frequency. The NASA EOSDIS Land Processes Distributed Active
Archive Center (LP DAAC), USGS Earth Resources Observation and Science
(EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and
MCD12Q1 from the online Data Pool, accessible at
https://lpdaa.usgs.gov/tools/data-pool/.

@@

To derive the drought indices of water supply and demand, soil moisture,
and vegetation (i.e., the proxy of productivity), we used the ERA5L
dataset and the MODIS product, with a monthly frequency for 1981--2023
and 2000--2023, respectively.

The dought indices correspond to a historical anomaly with regard to a
variable (e.g., meteorological, vegetation, or soil moisture). To
account for the anomaly, the common practice is to derive it following a
statistical parametric methodology in which it is assumed that the
statistical distribution of the data is known (\citet{Heim2002}). A
wrong decision is usually the highest source of uncertainty
(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
of climatic variability, it is complex to choose a proper distribution
without previous research. Here, we follow a non-parametric methodology
for the calculation of the drought indices, in a similar manner as the
framework proposed by \citet{Farahmand2015};
\citet{Hobbins2016};\citet{McEvoy2016}.

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which the World
Meteorological Organization (WMO) recommended. The SPI solely relies on
precipitation data. Also, it has been used worldwide for the study of
drought, including in Chile (\citet{Garreaud2017};
\citet{Zambrano2017}). The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse \citep{Luo2017}.
Nowadays, there is an increase in attention toward using water demand
separately to monitor droughts. (\citet{Vicente-Serrano2020};
\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
Evaporative Demand Drought Index (EDDI), developed by
\citet{Hobbins2016} and \citet{McEvoy2016}, which is based on the AED.
EDDI is currently used for monitoring drought in the United States
(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
AED, a difference from the original formulation of EDDI, which also
considered wind besides temperature. To consider the combined effect of
water supply and demand, we selected the SPEI, which corresponds to a
balance between precipitation and AED. \citet{Vicente-Serrano2010}
proposed the SPEI, and it has improved the SPI by incorporating
temperature for drought monitoring. For SPEI, an auxiliary variable D =
P-AED is calculated. Soil moisture is the main driver of vegetation
productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
for soil water drought, we used the SSI (Standardized Soil Moisture
Index) \citep{Hao2013, AghaKouchak2014} which is a multi-scale index
similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the
average soil moisture from ERA5L at 1m depth. Finally, for the proxy of
productivity, we used the zcNDVI proposed by \citet{Zambrano2018} which
will be derived from the NDVI retrieved from MOD13A1.

@@

The \(A_{si}\) corresponds to a moving window (convolution) that sums
the variable for time scales \(s\) from the last month, month by month,
until the first month in which it could sum for \(s\) months. Once the
variable is cumulated over time for the scale \(s\). Thus, the
empirically derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(a_i)\) probabilities across a period of interest:

@@

\(DI\) is referring to the drought index calculated for the variable
\(V\). The values for the constats are: \(C_0 = 2.515517\),
\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A) \leq 0.5\),
W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI) it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months; we choose to use six months
because that shows a more robust representation of vegetation
productivity due to the seasonality of vegetation in Chile.

@@

To analyze the land cover change, we use the IGBP scheme from the
MCD12Q1 collection 6.1 from MODIS. \citet{Zambrano2018} and
\citet{Fuentes2021} have previously used this product for studies of
drought and land cover in Chile. The MCD12Q1 has a yearly frequency from
2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into
ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands,
8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to
croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
water bodies. Thus, we have a land cover raster time series with the ten
macroclasses for 2001 and 2023. We validate the land cover macroclasses
regarding a highly detailed (30 m of spatial resolution) land cover map
made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
global accuracy of \textasciitilde0.82 and a F1 score of
\textasciitilde0.66. Section S2 in the Supplementary Material shows the
procedure for validation.

Climate, vegetation development, seasonality, and changes in vegetation
type all have an impact on the time series of NDVI. In this study, we
want to examine the variation in vegetation productivity across various
land cover types and how water demand, water supply, and soil moisture
affect it. In order to avoid changes due to a change in the land cover
type that will wrongly impact NDVI, we developed a persistence mask for
land cover for 2001--2022. Thereby, we reduce an important source of
variation on a regional scale. Therefore, we generated a raster mask for
IGBP MODIS per pixel using macroclasses that remained unchanged for at
least 80\% of the years (2001--2022). This enabled us to identify
regions where the land cover macroclasses are persistent. We calculated
the surface occupied per land cover class into the five macrozones
(``Norte Grande'' to ``Austral'') per year for 2001--2023. After that,
we calculated the trend's change in surface per type. We used the Sen'
slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.

@@

We wanted to explore the relationship between the trend in land cover
classes and the trend in the drought indices. For this purpose, in order
to have more representative results, we conducted the analysis over
sub-basins within continental Chile. We use 469 basins, which have a
surface area between 0.0746 and 24,000 (\(km^2\)), and a median area of
1,249 (\(km^2\)). For each basin, we calculate the relative trend per
land cover type, considering the proportion of the type relative to the
total surface of the basin. Then, we extracted per basin the average
trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time
scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in
the proxy of vegetation productivity (zcNDVI). We wanted to analyze
which drought indices and time scales have a major impact on changes in
land cover type.

We have 25 predictors, which are drought indices and vegetation
productivity. We analyzed the 25 predictors per type of landcover. For
the analysis, we selected the method of random forest (\citet{Ho1995}).
Because it allows to find no linear relationship, it reduces overfitting
and can derive the feature importance, which helps for a better
understanding of the relationships. The importance of the variable is
calculated by permuting out-of-bag (OOB) data per tree and calculating
the mean standard error in the OOB. Then the same is done after
permuting each predictor variable. Random forest uses multiple decision
trees and allows for classification and regression.

We analyzed the 25 predictors per type of landcover, thus running six
models. We used random forests for regression and trained 1000 forests.
For more reliable results for the important variables, we resampled by
creating ten folds, running a random forest per fold, and calculating
the r-squared (rsq), root mean square error (RMSE), and variable
importance ten times.

@@

We analyzed the trend of vegetation productivity over the unchanged land
cover macroclasses. This way, we tried to reduce the noise in the
vegetation due to a change in land cover from year to year. To achieve
this, we will use the persistent mask of land cover macroclasses, which
are the types that remain more than 80\% of the time for 2001--2022. We
used this to evaluate the trend in zcNDVI per land cover class and
macrozone.

We examine the drought indices of water demand, water supply, soil
moisture, and their connection with vegetation productivity to
investigate two main questions: i) whether short-term or long-term time
scales have a greater impact on vegetation across Chile and its specific
regions; and ii) the spatial variation in the strength of the
correlation between the variables and time scales.~ Then, we will
summarize for each land cover class and macrozone. Thus, we will be able
to advance in understanding how climate is affecting vegetation,
considering the impact on the five macroclasses of vegetation: forest,
cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months, and zcNDVI. The objective is to determine the impact of soil
moisture and water demand and supply on vegetation productivity. We used
a method similar to that used by \citet{Meroni2017} which compared the
SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically
Active Radiation). A pixel-to-pixel linear correlation analysis was
performed for each index within the persistent mask of land cover
macroclasses. To begin, the Pearson coefficient of correlation is
computed for each of the six time scales. A significant time scale is
identified as the one that attains the highest correlation (p
\textless{} 0.05). Subsequently, the Pearson correlation coefficient
corresponding to the time scales at which the value peaked was
extracted. As a result, for each index, we generated two raster maps: 1)
containing the raster with values of the time scales that reached the
maximum correlation, and 2) having the value of the correlation
obtained.

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
\{tidyverse\} \citep{Wickham2019} was used. For the random forest
modeling we used the \{tidymodels\}\citep{Kuhn2020} and
\{ranger\}\citep{Wright2017} packages.

@@

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
the macroclasses through Chile for the year 2022.
Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
persistance (80\%) during 2021--2022, respectively (Table
\ref{tab-landcoverSurf}). Within continental Chile, barren land is the
land cover class with the highest surface area (277,870 km2). The
largest type of vegetation, with 137,085 km2, is forest. Grassland
(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
(3,146 km2) are the other types (Table \ref{tab-landcoverSurf}). The
macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
and ``Austral,'' with 36\%, 31\%, and 34\% of their surface changing the
type of land cover, respectively (Figure~\ref{fig-studyArea} and Table
\ref{tab-landcoverTrend}). Figure~\ref{fig-LCprop} shows the summary of
the proportion of surface per land cover class and macrozone, derived
from the persistance mask over continental Chile.

@@

The ``Norte Chico'' shows an increase in barren land of 111
\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a shift in the area cultivated
(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (Table \ref{tab-landcoverTrend}).

@@

\hypertarget{vegetation-productivity}{%
\subsubsection{Vegetation productivity}\label{vegetation-productivity}}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend; reder
colors correspond to a negative trend and a decrease in vegetation
productivity. Grey colors indicate either no vegetation or a change in
land cover type for 2001--2022. (b) Temporal variation of zcNDVI-6
aggregated at macrozone level within continental Chile. Each horizontal
panel corresponds to a macrozone from `Norte Grande' to `Austral'.}

@@

In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
zcNDVI (Figure~\ref{fig-zcNDVI_var}a) and the temporal variation of
zcNDVI within the aggregated macrozones (Figure~\ref{fig-zcNDVI_var}b).
In ``Norte Grande,'' vegetation productivity, as per the z-index,
exhibits a yearly increase of 0.02 with respect to grassland and
shrubland categories. There is a negative trend in ``Norte Chico'' with
-0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
savanna (-0.05) has the lowest trend, and the rest of the types are
around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
and croplands and savanna -0.01 per decade. This could be associated
either with a reduction in vegetation surface, a decrease in biomass, or
browning \citep{Miranda2023}. Vegetation reached its lowest values since
the year 2019, with an extreme condition in early 2020 and 2022 in the
``Norte Chico'' and ``Centro''. The ``Sur'' and ``Austral'' show a
positive trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}).
Despite the croplands suffering from drought just as badly as the native
vegetation in ``Norte Chico,'' the savanna and shurbland appears to be
the region most affected by a negative trend in vegetation productivity.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

To analyze land cover change, we use the classification scheme by the
IGBP (International Geosphere-Biosphere Programme) from the product
MCD12Q1 collection 6.1 from MODIS. The MCD12Q1 has a yearly frequency
from 2001 to 2022 and defines 17 classes. To derive a proxy for
vegetation productivity, we used the Normalized Difference Vegetation
Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS
\citep{Didan2015}. MOD13A3 provides vegetation indices at 1km of spatial
resolution and monthly frequency. The NASA EOSDIS Land Processes
Distributed Active Archive Center (LP DAAC), USGS Earth Resources
Observation and Science (EROS) Center, Sioux Falls, South Dakota,
provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible
at https://lpdaa.usgs.gov/tools/data-pool/.

@@

To derive the drought indices of water supply and demand, soil moisture,
and vegetation (i.e., the proxy of productivity), we used the ERA5L
dataset and the MODIS product, with a monthly frequency for 1981--2023
and 2000--2023, respectively. The dought indices correspond to a
historical anomaly with regard to a variable (e.g., meteorological,
vegetation, or soil moisture). To account for the anomaly, the common
practice is to derive it following a statistical parametric methodology
in which it is assumed that the statistical distribution of the data is
known \citep{Heim2002}. A wrong decision is usually the highest source
of uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
high degree of climatic variability, it is complex to choose a proper
distribution without previous research. Here, we follow a non-parametric
methodology for the calculation of the drought indices, in a similar
manner as the framework proposed by \citet{Farahmand2015};
\citet{Hobbins2016};\citet{McEvoy2016}.

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which relies on
precipitation data. To evaluate water demand, we chose the Evaporative
Demand Drought Index (EDDI), developed by \citet{Hobbins2016} and
\citet{McEvoy2016}, which is based on the AED. The United States
currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/)
as an experimental index. To consider the combined effect of water
supply and demand, we selected the SPEI \citep{Vicente-Serrano2010}. For
SPEI, an auxiliary variable \(D = P-AED\) is calculated. Soil moisture
is the main driver of vegetation productivity, particularly in semi-arid
regions \citep{Li2022}. Hence, for soil water drought, we used the SSI
(Standardized Soil Moisture Index) \citep{Hao2013, AghaKouchak2014}. In
our case, for the SSI, we used the average soil moisture from ERA5L at
1m depth. Finally, for the proxy of productivity, we used the zcNDVI
proposed by \citet{Zambrano2018}, which was derived from the monthly
time series of NDVI retrieved from MOD13A1. All the indices are
multi-scalar and can be used for the analysis of short- to long-term
droughts.

@@

The \(A_{si}\) corresponds to a moving window (convolution) that sums
the variable for time scales \(s\) from the last month, month by month,
until the first month in which it could sum for \(s\) months. An inverse
normal approximation \citep{Abramowitz1968} obtains the empirically
derived probabilities once the variable cumulates over time for the
scale \(s\). Then, we used the empirical Tukey plotting position
\citep{Wilks2011} over \(A_i\) to derive the \(P(a_i)\) probabilities
across a period of interest:

@@

\(DI\) is referring to the drought index calculated for the variable
\(V\) (i.e., SPI, SPEI, EDDI, SSI, and zcNDVI). The values for the
constats are: \(C_0 = 2.515517\), \(C_1 = 0.802853\),
\(C_2 = 0.010328\), \(d_1 = 1.432788\), \(d_2 = 0.189269\), and
\(d3 = 0.001308\). For \(P(A) \leq 0.5\),
W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI) it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months; we choose to use six months
because that shows a more stable representation of vegetation
productivity.

@@

To analyze the land cover change, we use the IGBP scheme from the
MCD12Q1 collection 6.1 from MODIS. This product has been previously used
for studies of drought and land cover in Chile
\citep{Fuentes2021, Zambrano2018}. From the 17 classes, we regrouped
into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to
shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14
to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
water bodies. Thus, we have a land cover raster time series with the ten
macroclasses for 2001 and 2023. We validate the land cover macroclasses
regarding a highly detailed (30 m of spatial resolution) land cover map
made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
global accuracy of \textasciitilde0.82 and a F1 score of
\textasciitilde0.66. Section S2 in the Supplementary Material shows the
procedure for validation.

We calculated the surface occupied per land cover class into the five
macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
After that, we calculated the trend's change in surface per land cover
type and macroclass. We used Mann-Kendall for the significance of the
trend \citep{Kendall1975} and Sen's slope to calculate the magnitude
\citep{Sen1968}.

Later in this study, we will examine the variation in vegetation
productivity across various land cover types and how water demand and
supply, and soil moisture affect it. In order to avoid variations due to
a change in the land cover type from year-to-year that will wrongly
impact NDVI, we developed a persistence mask for land cover for
2001--2022. Thereby, we reduce an important source of variation on a
regional scale. Therefore, we generated a raster mask for IGBP MODIS per
pixel using macroclasses that remained unchanged for at least 80\% of
the years (2001--2022). This enabled us to identify regions where the
land cover macroclasses are persistent.

@@

We wanted to explore the relationship between the trend in land cover
classes and the trend in the drought indices. For this purpose, in order
to have more representative results, we conducted the analysis over
sub-basins within continental Chile. We use 469 basins, which have a
surface area between 0.0746 and ~24,000 (\(km^2\)), and a median area of
1,249 (\(km^2\)). For each basin, we calculate the relative trend per
land cover type, considering the proportion of the type relative to the
total surface of the basin. Then, we extracted per basin the average
trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time
scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in
the proxy of vegetation productivity (zcNDVI). We wanted to analyze
which drought indices and time scales have a major impact on changes in
land cover type.

We have 25 predictors, including drought indices and vegetation
productivity. We analyzed the 25 predictors per type of landcover, thus
running six random forest models. Random forest uses multiple decision
trees and allows for classification and regression. Some advantages are
that it allows to find no linear relationship, reduces overfitting, and
allows to derive the variable importance. We used random forests for
regression and trained 1000 forests. To obtain more reliable results, we
resampled by creating ten folds, running a random forest per fold, and
calculating the r-squared (rsq), root mean square error (RMSE), and
variable importance.

The variable importance helps for a better understanding of the
relationships by finding which variable has a higher contribution to the
model. We calculate the variable's importance by permuting out-of-bag
(OOB) data per tree and computing the mean standard error in the OOB.
After permuting each predictor variable, we repeat the process for the
resting variable. We repeated this process ten times (per fold) to
obtain the performance metrics (rsq, RMSE, and variable importance).

@@

We analyzed the trend of vegetation productivity over the unchanged land
cover macroclasses. To achieve this, we used the persistent mask of land
cover macroclasses. This way, we tried to reduce the noise in the
vegetation due to a change in land cover from year to year.

We examine the drought indices of water demand, water supply, and soil
moisture and their correlation with vegetation productivity. The
objective is to determine the impact of soil moisture and water demand
and supply on vegetation productivity. We want to address three main
questions: Which of the drought variables---supply, demand, or soil
moisture---most helps to explain the changes in plant productivity?
Which of the short-term or long-term time scales of the drought variable
has a bigger effect on vegetation productivity in Chile? And finally,
how strong is the relationship between the variables and the drought
index? Thus, we will be able to advance in understanding how climate is
affecting vegetation, considering the impact on the five land cover
types: forest, cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months with zcNDVI. We used a method similar to that used by
\citet{Meroni2017} which compared the SPI with the cumulative FAPAR
(Fraction of Absorbed Photosynthetically Active Radiation). A
pixel-to-pixel linear correlation analysis was performed for each index
within the persistent mask of land cover macroclasses. To begin, the
Pearson coefficient of correlation is computed for each of the six time
scales. A time scale is identified as the one that attains the highest
correlation (p \textless{} 0.05). Subsequently, the Pearson correlation
coefficient corresponding to the time scales at which the value peaked
was extracted. As a result, for each index, we generated two raster
maps: 1) containing the raster with values of the time scales and
drought index that reached the maximum correlation, and 2) having the
magnitude of the correlation obtained by the drought index at the time
scales.

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
\{tidyverse\} \citep{Wickham2019} was used. For the random forest
modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
\{ranger\}\citep{Wright2017} packages.

@@

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
the macroclasses through Chile for the year 2022.
Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
persistance (80\%) during 2021--2022, respectively (Table
\ref{tab-landcoverSurf}). Within continental Chile, barren land is the
land cover class with the highest surface area (277,870 \(km^2\)). The
largest type of vegetation, with 137,085 \(km^2\), is forest. Grassland
has 74,247 \(km^2\), savanna 55,206 \(km^2\), shrubland 25,341 \(km^2\),
and cropland 3,146 \(km^2\) (Table \ref{tab-landcoverSurf}). The
macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
and ``Austral,'' with 36\%, 31\%, and 34\% of their surface changing the
type of land cover, respectively (Figure~\ref{fig-studyArea} and Table
\ref{tab-landcoverTrend}). Figure~\ref{fig-LCprop} shows the summary of
the proportion of surface per land cover class and macrozone, derived
from the persistance mask over continental Chile.

@@

From the trend analysis in Table \ref{tab-landcoverTrend}, we can
indicate that the ``Norte Chico'' shows an increase in barren land of
111 \(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. The area under cultivation (croplands) appears to be
shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (Table \ref{tab-landcoverTrend}).

@@

\hypertarget{trends-in-vegetation-productivity}{%
\subsubsection{Trends in vegetation
productivity}\label{trends-in-vegetation-productivity}}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI for 2000--2023. Greener colors indicate a positive trend; reder
colors correspond to a negative trend and a decrease in vegetation
productivity. Grey colors indicate either no vegetation or a change in
land cover type for 2001--2022. (b) Temporal variation of zcNDVI-6
aggregated at macrozone level within continental Chile. Each horizontal
panel corresponds to a macrozone from `Norte Grande' to `Austral'.}

@@

In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
zcNDVI (Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande,'' vegetation
productivity, as per the z-index, exhibits a yearly increase of 0.027
for grassland and 0.032 for shrubland. In the ``Norte Chico,'' savanna
has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and
grassland -0.037. In ``Centro,'' shrubland reaches -0.07, savanna
-0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
This decrease in productivity could be associated either with a
reduction in vegetation surface, a decrease in biomass, or browning
\citep{Miranda2023}.

The temporal variation within the macrozones is showed in
Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
reached its lowest values since the year 2019, with an extreme condition
in early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The
``Sur'' and ``Austral'' show a positive trend of around 0.012 and 0.016,
respectively per decade (Figure~\ref{fig-zcNDVI_var}). Despite the
croplands suffering from drought just as badly as the native vegetation
in ``Norte Chico,'' the savanna and shurbland appears to be the region
most affected by a negative trend in vegetation productivity across
Chile.

### 2024-03-20 article/supplementary.qmd
update article

---
title: "Supplementary Material"
author: "Francisco Zambrano"
format: 
  pdf:
    number-sections: true
editor: source
---

@@

>>>>

---
title: "Supplementary Material"
author: "Francisco Zambrano"
date: "`r Sys.Date()`"
format: 
  pdf:
    number-sections: true
editor: source
---

@@

# Trend of vegetation productivity

```{r}
#| echo: false
#| out-width: 75%
#| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
```
